---
title: "R Basics & Data Cleaning"
author: "Kevin da Silva Castanheira"
date: "9/18/2018"
output: word_document
---

```{r setup, include=FALSE}

```

# R Basics

#### Asking for Help
If you ever get stuck and need more information on any function you can type the following commands into the console. They are equivalent.
```{r,results='hide'}
help(aggregate) 
?aggregate
```
#### Variable types

In R as with many other languages your variable can take on several different types:
Character: a combination of letters
```{r}
"Hi everyone"
```

Numeric: Number (integer or decimal)
```{r}
12
123.232412
```

Factor: A set of variables with pre-defined values
```{r}
factor(c("male", "female", "other","male", "female","male", "female","male", "female"))
factor(c("monday", "tuesday", "wednesday","thursday", "friday","saturday", "sunday"))
factor(c(1,0,1,1,0,1,0,1))
```


#### Vectors, Data-frames, Lists

R has a few basic data structures:

##### Vector: a combiantion of variables with the same type
```{r}
c(1,2,3,4,5)
5:10
c("Dog", "Cat", "Bird", "Fish")
```

##### List: a combiantion of variables with no restriction on type
```{r}
list(c(1,2,3,4,5), c("Dog", "Cat", "Bird", "Fish"), c(TRUE, FALSE))
```

##### Data Frame: a matrix organizing data in rows and columns
```{r}
data.frame("PID"=1:3, "IQ"=c(100, 130, 90), "Age"=c(24, 23, 21))
```

#### Unless you save your variable by assigning it a "name", the output will not be retrievable to manipulate
```{r}
df = data.frame("PID"=1:3, "IQ"=c(100, 130, 90), "Age"=c(24, 23, 21))
df[1,2] = 40
df
```


## Boolean/Logical variables

In most programming languages, there exists boolean/logical variables. They take on the value of TRUE or FALSE. In R as with other languages you can apply operators on them.
```{r}
1 > 3
3 > 1
3.1 >= 1
3.1 <= 1
```
You can test the equality of two things by using double equal signs.
```{r}
3 == 1
"Cat" == "Cat"
"CAT" == "Cat"
```
You can also test if a value or multiple values is/are present in a vactor:
```{r}
1 %in% c(2,3,4,5)
c(1,2,3) %in% c(2,3,4,5)
```

##### Boolean Operators
In R, the ! symbolizes the NOT operator. It will return the opposite value of the boolean.
```{r}
3 != 1
!TRUE
!FALSE
```

You can combine boolean variables by using the AND/OR operators.

AND requires both variables to be TRUE to return TRUE
```{r}
(1 > 4) & (1 < 4)
(6 > 4) & (1 < 4)
```
OR only requires that one variable equal to TRUE
```{r}
(1 > 4) | (1 < 4)
```


You can also use if and else statements to control the output of your code.
#### If/Else statements
```{r}
if(1 < 2){
  print("TRUE")
}

if(3 < 1){
  print("IMPOSSIBLE")
} else{
  print("3 is greater than 1")
}

if(3 < 1){
  print("IMPOSSIBLE")
} else if(3 > 1){
  print("3 is greater than 1")
} else{
  print("WHAT?")
}

```

## For loops
If you need to do something multiple time you can use loops to accomplish this.

For loops work by repeating whatever is inside them until a pre-specified condition is met.
```{r}
for(i in 1:7){
  print(i)
}

```

While loops are similar except function without needing to specify exactly how many time to repeat a process.
```{r}
t = 0
while(t<0.5){
  t = runif(1,0,1) # change the value of t by sampling from a uniform distribution
  # the first argument into the function specifies the number of sample
  # the second and third arguments specify the min and max of the range respectively
  print(t)
}

```

# Loading Data
To load data you first need to know a few things about paths.

1) You do not need to provide a full path if the file you're interested in is in the current working directory

```{r}
getwd()
```
If this is not where you want your working directory to point you can change it
```{r, warning=FALSE, message=FALSE}
setwd("~/Desktop")
```

2) If you are on a Mac or a Linux OS paths look different than on a PC
MAC/LINUX: "/Users/kevin/Desktop/File" OR "~/Desktop/File"
PC: "C:\Users\Kevin\Desktop\File"
You can load data into R from a csv using the command read.csv. 
```{r}
NFC <- read.csv("~/Coding-Club/R_Basics/NFC.csv")

```
If the first column contains row names, you can specify this to the function in order not read it in as a new column.
```{r}
df <-read.csv("~/Coding-Club/R_Basics/data.csv", sep = ',',
             row.names = 1)
```

OR if your data is stored in some other format (separated by semi-colons) and does not contain headers, you can modify the arguments into the function read.csv to read the data properly.
```{r}
Digit_Symbol <- read.csv("~/Coding-Club/R_Basics/DigitSymbol.txt", header=FALSE, sep=";")

colnames(Digit_Symbol) = c("PID", "Score")
```

#### Looking at your data
Get thr first 10 values
```{r}
head(NFC, 10)
```
Or the last 10
```{r}
tail(NFC,10)
```

Or a summary of all the columns
```{r}
summary(NFC)
```
Or you can look at a single column by indexing into the data frame
```{r, results='hide'}
NFC$PID
NFC[,1] # format is DATAFRAME[ROW, COLUMN]
```

You can apply various functions to your columns. One useful function is unique which will return all the unique values of a column. I used it here to get a list of all Participant IDs.
```{r}
unique(NFC$PID)
```

# Data Cleaning
There are many useful function which can allow you to look at and manipulate you data.
One useful feature of R is to remove trials or participants.

is.na is a function which checks if a value is NA.
Here we are specifying we only want the rows in which Gender is not NA.
```{r}
NFC = NFC[!is.na(NFC$Gender),]
```

Or we can exclude participants with an IQ score below a certain threshold
```{r}
NFC = NFC[NFC$IQ > 80,]
```

You can index into a data frame to change the values.
For example, here I am reverse coding the values of certain questions.
```{r}
NFC_reversal = c("Q3", "Q4", "Q5", "Q7", "Q8", "Q9", "Q12", "Q16", "Q17")
NFC$Value[NFC$Question %in% NFC_reversal] = 6 - NFC$Value[NFC$Question %in% NFC_reversal]
```

#### GREP
You can use regular expressions to find patterns of characters and numbers. In our example, we can try to classify people based on their highest level of education...
```{r}
unique(NFC$Education)
```
Some rules of GREP 
(see the link for more info: https://stat.ethz.ch/R-manual/R-devel/library/base/html/regex.html)

1) It is case sensitive Case so make sure everything is in the same case
```{r}
NFC$Education = toupper(NFC$Education)
unique(NFC$Education)
```

2) things preceeded by * are matched 0 or more times
3) things preceeded by ? will be matched 0 or 1 times
4) ^ matches the start of the string of characters
5) $ matches the start of the end of characters
```{r}

NFC$Bachelors = grepl("B[.]?S[C]?[.]?[A]?[ACHELOR]?[S]?", NFC$Education) | grepl("UNDERGRAD[UATE]?", NFC$Education)

NFC$HS = NFC$Education == "HIGH SCHOOL"

NFC$Grad = grepl("^GRAD", NFC$Education) | grepl("M[.]?[A]?[S]?[SC]?[.]?", NFC$Education) | grepl("PH[.]?D", NFC$Education)
```

# Pivot Tables in R
If you're familiar with Excel, then the function aggregate will be very familiar.
The function aggregate allows you to apply a function to groups you define. For example, if we want to calculate a score per participant you can sum all the value for every participant.

The variable on the left of the ~ is the "outcome variable" you are applying the function to. The variables on the right of the ~ are the grouping variables. The final two inputs are the data frame and the function respectively
```{r}
Scores = aggregate(Value ~ PID, NFC, sum)
head(Scores, 10)
```
You can add several columns to this.
```{r}
aggregate(Age ~ Gender+Bachelors, NFC, mean)
```
And you can even define your own function! Here I'm defining my own function (SEM).
```{r}
aggregate(Age ~ Gender+Bachelors, NFC, function(x) sd(x)/sqrt(length(x)))
```

# Mapping Values Between data-frames
Mapvalues allows you to map data from a different data frame and map those values to another data frame based on certain values. For example, mapping the scores of participants from one data frame of a questionnaire to another data frame.

mapvalues takes in 3 arguments

1) the factor to modify
2) the original values (which match those in 1)
3) the new values you want to map

since mapvalues is a function not availble in base R, it is from the library plyr, you either need to laod the library plyr or let R know the function is from that library by putting its name before the function name.
```{r}
NFC$DigitSymbol = plyr::mapvalues(NFC$PID, Digit_Symbol$PID, Digit_Symbol$Score)
```
