---
title: "The ggplot Bible"
author: "Kevin da Silva Castanheira"
date: "August 14, 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting Started
### Start by loading the Library dplyr, ggplot2, grid, and gridExtra

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(grid)
library(gridExtra)
Quebec = read.csv("~/coding-club/advanced_ggplot/Quebec.csv")
```
# ggplot & dplyr Basics

Let us refresh ourselves with the basics of dplyr and ggplot.

## dplyr

You can use dplyr to manipulate your data and make new columns. Here we will focus on summarizing data (i.e. data reduction).

Let's take a look at the dataframe Quebec. It has 26 columns and a bunch of rows reflecting restaurants in the greater montreal area and their corresponding Yelp reviews. 

```{r, messages=FALSE, warning=FALSE}
head(Quebec, 2)
summary(Quebec)
```

Now let's try to get a mean rating for each postal code. The functions group_by will split your data into groups, piping (%>%) that to the function summarise will produce a summary variable.

```{r, messages=FALSE, warning=FALSE}
Quebec %>% group_by(postal_code) %>% summarise(Mean = mean(stars))
```

You can get mode than one outcome variable in summarise.

```{r, messages=FALSE, warning=FALSE}
Quebec %>% group_by(postal_code) %>% summarise(Mean = mean(stars), SD= sd(stars), N = n())
```

And you can choose more than one variable to group by.

```{r, messages=FALSE, warning=FALSE}
Quebec %>% group_by(postal_code, PriceRange) %>% summarise(Mean = mean(stars), SD= sd(stars), N = n())
```

## ggplot

ggplot works by taking your data and plotting it. I usually like to summarise my data and then pipe it to ggplot.

Every plot starts with the base ggplot statement. This is where you will tell R how you want your graph to be organized (what is your x, y, color variable etc.)

THIS ITSELF WONT PRODUCE A PLOT!!!!

You need to tell R what "shapes" you want it to draw. This is where geom comes in. 

For example you can draw bar graphs:

```{r, messages=FALSE, warning=FALSE}
Quebec %>% group_by(PriceRange) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars)) + geom_bar(stat = "summary", position="dodge") + theme_minimal()
```

For example you can draw scatter plots:

```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=city, size=city)) + geom_point(alpha=0.5) + theme_minimal()
```

For example you can draw histograms:

```{r, message=FALSE}
Quebec %>% 
  ggplot(aes(stars)) + geom_histogram() + theme_minimal()
```

For example you can draw Violin plots:

```{r, messages=FALSE, warning=FALSE}
Quebec %>% 
  ggplot(aes(Portuguese, stars)) + geom_violin() + theme_minimal()
```
... etc.

You can add geoms (error bars, best fit lines etc.) to these basic plots to create more complex plots.

```{r, messages=FALSE, warning=FALSE}
Quebec %>% group_by(PriceRange) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars)) + geom_bar(stat = "summary", position="dodge") + geom_errorbar(aes(ymin=stars-sem, ymax=stars+sem), width=0, position="dodge") + theme_minimal()
```

#### Best Fit lines
Best fit lines are added using the geom_smooth() function.
geom_smooth() allows you to fit many kinds of fucntions but for now we will explore the linear fits.
```{r, messages=FALSE, warning=FALSE}
Quebec %>% group_by(PriceRange) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars)) + geom_point() + geom_errorbar(aes(ymin=stars-sem, ymax=stars+sem), width=0) + geom_smooth(method="lm") + theme_minimal()
```

And you can remove the ugly shaded grey area (standard error) by supplying the function geom_smooth with the argument se=FLASE.
```{r, messages=FALSE, warning=FALSE}
Quebec %>% group_by(PriceRange) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars)) + geom_point() + geom_errorbar(aes(ymin=stars-sem, ymax=stars+sem), width=0) + geom_smooth(method="lm", se=FALSE) + theme_minimal()
```

Specifying a color/fill/grouping variable will result in two lines fit--one for each group.
```{r, messages=FALSE, warning=FALSE}
Quebec %>% group_by(PriceRange, Italian) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars, color=Italian)) + geom_point() + geom_errorbar(aes(ymin=stars-sem, ymax=stars+sem), width=0) + geom_smooth(method="lm", se=FALSE) + theme_minimal()
```
### Finishing touches
#### Color, Size, alpha
You can change the color, fill, size or alpha of your plot based on a variable in your dataset.

Setting color by a continuous (numeric) varibale will make a gradient
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() + theme_minimal()
```

Which you can change:
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() + scale_color_gradientn(colours=c("deepskyblue", "hotpink1")) + theme_minimal()
```
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() + scale_color_gradientn(colours=c("deepskyblue", "white", "hotpink1")) + theme_minimal()
```

Or you can map color using discrete variables and not make a gradient:
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=city)) + geom_point() + theme_minimal()
```

And you can change their values also:
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=city)) + geom_point() + scale_color_manual(values=c("deepskyblue", "hotpink1", "orchid")) + theme_minimal()
```
For example here we have color and size mapping to city and opacity (alpha) set to 0.5:

```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=city, size=city)) + geom_point(alpha=0.5) + theme_minimal()
```

And here we have color and alpha mapping to city and size set to 2:
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=city, alpha=city)) + geom_point(size=2) + theme_minimal()
```
You can even have color, alpha, size mapped to different variables:

```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=city, size=French)) + geom_point(alpha=0.5) + theme_minimal()
```

#### Title & Labels
You can title your graph by using the function ggtitle() and label you x and y axis with the functions xlab() and ylab()

```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() +ggtitle("TITLE") + xlab("X") + ylab("Y") 
```
You can even change the label of the legend:
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=city, size=city)) + geom_point() +ggtitle("TITLE") + xlab("X") + ylab("Y") + labs(color="TEST 1", size="TEST 2")
```
#### Theme
You can select a theme to change the look of your graph:
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() 
```

```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() + theme_classic()
```

```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() + theme_void()
```
You can also manually change these values yourself to your liking including font size, centering the title, etc. 

Center the title:
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() +
  ggtitle("TITLE") + theme(plot.title = element_text(hjust = 0.5))
```

Remove the Grid lines:
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() +
  ggtitle("TITLE") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

Remove the background color:
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() +
  ggtitle("TITLE") + theme(panel.background = element_blank())
```

Change the border of the panels
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() +
  ggtitle("TITLE") + theme( panel.border = element_rect(colour = "black", fill=NA, size=0.5))
```

Or change the font size
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() +
  ggtitle("TITLE") + theme(text = element_text(size=20))
```

Or do all of it together...
```{r, message=FALSE, warning=FALSE}
Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=PriceRange)) + geom_point() +
  ggtitle("TITLE") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(),  panel.border = element_rect(colour = "black", fill=NA, size=0.5),
        legend.key.size = unit(1,"cm"), text = element_text(size=20), 
        plot.title = element_text(hjust = 0.5))
```


#### Horizontal and Verticle Lines
You can add lines onto your graphs to demark important values.

hline adds horizontal lines (i.e. to demark important y values like p=0.5 if plotting probabilities):

```{r, messages=FALSE, warning=FALSE}
Quebec %>% group_by(PriceRange) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars)) + geom_bar(stat = "summary", position="dodge") + geom_errorbar(aes(ymin=stars-sem, ymax=stars+sem), width=0, position="dodge") + theme_minimal() + geom_hline(yintercept=3)
```

And you can change the style of the line by specifying what kind of line you want (or a number 1 = solid, 2 = dashed, 3 = dotted, 4 = dotdash, 5 = longdash)
```{r, messages=FALSE, warning=FALSE}
Quebec %>% group_by(PriceRange) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars)) + geom_bar(stat = "summary", position="dodge") + geom_errorbar(aes(ymin=stars-sem, ymax=stars+sem), width=0, position="dodge") + theme_minimal() + geom_hline(yintercept=3, linetype='dashed')
```

Or you can go crazy and add multiple!!!

```{r, messages=FALSE, warning=FALSE}
Quebec %>% group_by(PriceRange) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars)) + geom_bar(stat = "summary", position="dodge") + geom_errorbar(aes(ymin=stars-sem, ymax=stars+sem), width=0, position="dodge") + theme_minimal() + geom_hline(yintercept=seq(0.5,4.5, 0.5), linetype=4)
```

vline adds a verticle line (to demark important x values like stimulus onset):

```{r, messages=FALSE, warning=FALSE}
Quebec %>% 
  ggplot(aes(review_count, stars)) + geom_point() + theme_minimal() + geom_vline(xintercept=0, linetype=3)
```

```{r, messages=FALSE, warning=FALSE}
Quebec %>% 
  ggplot(aes(review_count, stars)) + geom_point() + theme_minimal() + geom_vline(xintercept=seq(0, 15, 2.5), linetype=5)
```

# Facet Wrapping

Facet wrapping is a way of separating your graph into panels based on variables within your dataset. 

```{r, messages=FALSE, warning=FALSE}
Quebec %>%
  ggplot(aes(review_count, stars)) + geom_point(stat = "summary", position="dodge")  + theme_minimal() + facet_wrap(~PriceRange)
```

You can specify how you want these plots to be arranged within your graph by specifying the number of columns or rows.

```{r, messages=FALSE, warning=FALSE}
Quebec %>%
  ggplot(aes(review_count, stars)) + geom_point(stat = "summary", position="dodge")  + theme_minimal() + facet_wrap(~PriceRange, nrow=1)
```
```{r, messages=FALSE, warning=FALSE}
Quebec %>% 
  ggplot(aes(review_count, stars)) + geom_point(stat = "summary", position="dodge")  + theme_minimal() + facet_wrap(~PriceRange, ncol=1)
```

You can facet wrap by multiple variables. The order of the variables you specify to facet wrap will change how the plot is made

```{r, messages=FALSE, warning=FALSE}
Quebec %>% 
  ggplot(aes(review_count, stars)) + geom_point(stat = "summary", position="dodge")  + theme_minimal() + facet_wrap(~PriceRange+French, nrow=2, ncol=4)
```
```{r, messages=FALSE, warning=FALSE}
Quebec %>% 
  ggplot(aes(review_count, stars)) + geom_point(stat = "summary", position="dodge")  + theme_minimal() + facet_wrap(~French+PriceRange, nrow=2, ncol=4)
```

You can pipe your data into functions like revalue form the plyr package (DO NOT LOAD THE PLYR PACKAGE AND DPLYR SIMULTANEOUSLY!!!!) to relabel your variables before graphing.

```{r, messages=FALSE, warning=FALSE}
Quebec$French = plyr::revalue(as.factor(Quebec$French), c("TRUE"="French", "FALSE"="Other"))
Quebec %>%
  ggplot(aes(review_count, stars)) + geom_point(stat = "summary", position="dodge")  + theme_minimal() + facet_wrap(~French+PriceRange, nrow=2, ncol=4)
```

Facet Wrapping is very useful when you want to arrange multiple plots with the same dependant and independant variables.

But what happens when we want to plot multiple graphs in the same figure with different variables?

# Arranging Plots

With the libraries grid and gridExtra, you can arrange plots of different variables all within the same figure

```{r, messages=FALSE, warning=FALSE}
plot1 <- Quebec %>% filter(city %in% c("MONTREAL", "LAVAL", "WESTMOUNT")) %>%
  ggplot(aes(PriceRange, stars, color=city, size=city)) + geom_point(alpha=0.5) + theme_minimal()

plot2 <- Quebec %>% group_by(PriceRange) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars)) + geom_bar(stat = "summary", position="dodge") + geom_errorbar(aes(ymin=stars-sem, ymax=stars+sem), width=0, position="dodge") + theme_minimal() + geom_hline(yintercept=3)

plot3 <- Quebec %>% 
  ggplot(aes(stars)) + geom_histogram() + theme_minimal()

plot4 <- Quebec %>% 
  ggplot(aes(French, stars)) + geom_violin() + theme_minimal()

grid.arrange(plot1, plot2, plot3, plot4)
```

You can change the way the plots are arranged by specifying the number of rows and columns
```{r, messages=FALSE, warning=FALSE}
grid.arrange(plot2, plot3, plot4, nrow=1)
grid.arrange(plot2, plot3, ncol=1, nrow=2)
```

You can achieve more complex designs (non-grid layouts) by specifying the widths and layouts you want.

Let's explore our options.

Let's make a 2x2 figure where the left graphs are twice as wide as the right hand graphs. Width specifies the width of each column. Layout_matrix specifies how the graphs are to be laid out, row-by-row.

```{r, messages=FALSE, warnings=FALSE}
grid.arrange(plot1, plot2, plot3, plot4,
  widths = c(2, 1),
  layout_matrix = rbind(c(1, 2),
                        c(3, 4)))

```

Now Let's make only the top-left graph large and the others small. Notice how we have to leave a blank space on the bottom row for the graph that is missing and the first graph takes up two columns.

```{r, messages=FALSE, warnings=FALSE}
grid.arrange(plot1, plot2, plot3, plot4,
  widths = c(1, 1, 1),
  layout_matrix = rbind(c(1, 1, 2),
                        c(NA, 3, 4)))

```

Now try making the top left graph be 3 times as large as the other graph in it's row but have the bottom two split evenly. FOr this we're going to need 6 columns.

```{r, messages=FALSE, warnings=FALSE}
grid.arrange(plot1, plot2, plot3, plot4,
  widths = c(1, 1, 1, 1, 1, 1),
  layout_matrix = rbind(c(1, 1, 1, 1, 2, 2),
                        c(3, 3, 3, 4, 4, 4)))

```

Now let's add a titles to the subplots!

```{r, messages=FALSE, warnings=FALSE}
grid.arrange(plot1 + ggtitle("A"), plot2 + ggtitle("B"), plot3 + ggtitle("C"), plot4 + ggtitle("D"),
  widths = c(1, 1, 1, 1, 1, 1),
  layout_matrix = rbind(c(1, 1, 1, 1, 2, 2),
                        c(3, 3, 3, 4, 4, 4)))

```

#### Titles 
What about an general title.

```{r, messages=FALSE, warnings=FALSE}
grid.arrange(plot1 + ggtitle("A"), plot2 + ggtitle("B"), plot3 + ggtitle("C"), plot4 + ggtitle("D"),
top=textGrob("Figure 1"),
  widths = c(1, 1, 1, 1, 1, 1),
  layout_matrix = rbind(c(1, 1, 1, 1, 2, 2),
                        c(3, 3, 3, 4, 4, 4)))

```

#### Graph Inserts

To insert a smaller graph in your other graph you first must turn your ggplot graph into a ggplorGrob. This will allow you to insert it into any figure.

```{r, messages=FALSE, warning=FALSE}
plot2 <- Quebec %>% group_by(PriceRange) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars)) + geom_bar(stat = "summary", position="dodge") + geom_errorbar(aes(ymin=stars-sem, ymax=stars+sem), width=0, position="dodge") + theme_minimal() + geom_hline(yintercept=3)

plot2Grob = ggplotGrob(plot2 + theme(plot.background = element_rect(colour = "black")))

Quebec %>% 
  ggplot(aes(stars)) + geom_histogram() + theme_minimal() +
  annotation_custom(
    grob = plot2Grob,
    xmin = 1,
    xmax = 3,
    ymin = 150,
    ymax = 300)

```

You can insert text.

```{r, messages=FALSE, warning=FALSE}
plot2 <- Quebec %>% group_by(PriceRange) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars)) + geom_bar(stat = "summary", position="dodge") + geom_errorbar(aes(ymin=stars-sem, ymax=stars+sem), width=0, position="dodge") + theme_minimal() + geom_hline(yintercept=3)

Grob_insert = textGrob("Kevin Made This!!")

Quebec %>% 
  ggplot(aes(stars)) + geom_histogram() + theme_minimal() +
  annotation_custom(
    grob = Grob_insert,
    xmin = 1,
    xmax = 3,
    ymin = 150,
    ymax = 300)

```

Or you can add a table which summarizes the average rating of Portuguese and non-Portuguese Restaurants.

```{r, messages=FALSE, warning=FALSE}
plot2 <- Quebec %>% group_by(PriceRange) %>% summarise(sem= sd(stars)/sqrt(length(stars)), stars=mean(stars)) %>% 
  ggplot(aes(PriceRange, stars)) + geom_bar(stat = "summary", position="dodge") + geom_errorbar(aes(ymin=stars-sem, ymax=stars+sem), width=0, position="dodge") + theme_minimal() + geom_hline(yintercept=3)

Grob_insert = tableGrob(Quebec %>% group_by(Portuguese) %>% summarise(mean=mean(stars)), rows = c())

Quebec %>% 
  ggplot(aes(stars)) + geom_histogram() + theme_minimal() +
  annotation_custom(
    grob = Grob_insert,
    xmin = 1,
    xmax = 3,
    ymin = 150,
    ymax = 300)

```

